---
title: "Extract and summarize length frequencies for Swordfish (Xiphias gladius)"
subtitle: "Submission to ISC BILLWG Jun 2017"
author: "M Kapur maia.kapur@noaa.gov"
output: 
  html_notebook:
    toc: yes
    toc_depth: 3
---

<style type="text/css">

body{ /* Normal  */
   font-size: 12px;
}
td {  /* Table  */
   font-size: 8px;
}
h1 { /* Header 1 */
 font-size: 18px;
 color: DARKCYAN;
}
h2 { /* Header 2 */
 font-size: 16px;
 color: LIGHTSEAGREEN;
}
h3 { /* Header 3 */
 font-size: 14px;
 color: DARKOLIVEGREEN;
}
code.r{ /* Code block */
  font-size: 12px;
}
pre { /* Code block */
  font-size: 10px
}
</style>
The purpose of this memorandum is to provide a collation of best available science regarding the life history & stock status of Swordfish (<i>Xiphias gladius</i>)  caught in the Hawaiian longline fishery.
What distinguishes the download protocol here is we aren't interested in the entire # of caught fish (wherein we would use the ORADATA schema); instead we want length measurements and set-level information, which requires use of the NEWOBS schema.<br>
<b> Message from E Forney 24 Mar:</b><br>
Hi Maia,

The reason only ~92k records are extracted since 2003 is that prior to August 2003 the observer program used different data forms. The current forms allow for measurements to be recorded directly on the Catch form with a species such as swordfish. Prior to August 2003 there was a separate Fish Life History form where observers recorded measurements for billfish, sharks and tuna. These measurements were not linked to particular billfish, sharks and tuna on the Catch form.

Also prior to about February 2006 our protocol was for observers to measure every billfish, tuna and shark that came onboard if it was safe to do so. In February 2006 we switched to having our observers measure every third fish regardless of species.

Running the script extracts over 250k swordfish records. However, the first 26k swordfish records from 1994 through August 2003 have no measurements for the aforementioned reasons.
select 
trip_num,
set_num,
species_code,
english_name,
haulbegin_day,
haulbegin_mon,
haulbegin_yr,
eye_fork
from newobs.lds_catch_v
where species_code = '091'
order by trip_num

Hope this helps.
Load Packages

```{r, warning = F, message = F, echo = F, eval = F}
## this is a dataframe Jon provided from B Walsh which represents the ideal summary format for length frequency data. It was used as a template to generate new data.
priorcatch = read.csv("G:/ISC BILLWG/length-freq/091_1994_2012.csv")
with(subset(priorcatch, HAULBEGIN_YR == 2012), max(HAULBEGIN_MON)) ## prior catch goes through oct 2012
priorcatch = priorcatch[,-1]
ldssum = read.csv('G:/ISC BILLWG/summary_LDS_count.csv')
with(ldssum, plot(as.numeric(HAULBEGIN_YR), COUNT))
```


## Acquire data
Using protocol from M Kapur data prep, 2017. Swordfish appear as CODE 6 in observer dataset, CODE 91 in catch dataset. Columns needed are spread out across multiple tables, namely because observer data before 2003 has the length records in a separate place LOP_BILLFISH_LEGACY_V.
[! B Walsh's instructions for length data per time period, TM-16-002a](G:/_LLDP/TM216_length.png)
```{r, warning = F, message = F, echo = F}
## load packages
library(RODBC)
library(reshape2)
library(plyr)
library(dplyr)
library(tidyverse)

## designate username and password
UID = c('UID','USERNAME')[2]
pwd = c('pwd', 'PASSWORD')[2]
con  =  odbcConnect(dsn="PIC", uid=UID, pwd = pwd, believeNRows=FALSE)
## use these tables to get specimen, set and trip level information. it helps to inspect which fields are available in each column before downloading.
# sqlTables(con, schema="NEWOBS")
# sort(sqlColumns(con, schema="NEWOBS",sqtable="LDS_CATCH_V")[,'COLUMN_NAME'])
# sort(sqlColumns(con, schema="NEWOBS",sqtable="LDS_SET_ENVIRON_V")[,'COLUMN_NAME'])
# sort(sqlColumns(con, schema="NEWOBS",sqtable="LDS_GEAR_CFG_V")[,'COLUMN_NAME'])
# sort(sqlColumns(con, schema="NEWOBS",sqtable="LDS_TRIPS_V")[,'COLUMN_NAME'])
sort(sqlColumns(con, schema="NEWOBS",sqtable="LDS_BILLFISH_LEGACY_V")[,'COLUMN_NAME'])
```


```{r}
## Longline Data
p = proc.time() ## less than 1 min




## legacy catch data
#!-- leg0 = sqlQuery(con, 
#"select 
#trip_num,
#set_num,
#species_code,
#species_common_name,
#haulbegin_day,
#haulbegin_mon,
#haulbegin_yr,
#gender as gender_code,
#eye_fork_len as eye_fork
#from newobs.LDS_BILLFISH_LEGACY_V
#and eye_fork_len is not null
#and eye_fork_len > '0'
#order by haulbegin_yr") -->

set = dbGetQuery(con, "SELECT 
TRIP_NUM,
SET_NUM,
HAULBEGIN_DAY,
HAULBEGIN_MON,
HAULBEGIN_YR,
HAUL_BEGIN_LAT,
HAUL_BEGIN_LON
from NEWOBS.LDS_SET_ENVIRON_V") 


gear = dbGetQuery(con, "SELECT 
TRIP_NUM,
SET_NUM,
HAULBEGIN_DAY,
HAULBEGIN_MON,
HAULBEGIN_YR,
HKS_PER_FLT,
TARGET_SPECIES_CODE,
TARGET_COMMON_NAME AS TARGET_SPECIES_NAME
from NEWOBS.LDS_GEAR_CFG_V")


## take out catch data from 2003 onwards (EYE_FORK values should start in August, )
catch0 =  subset(dbGetQuery(con, 
"select 
trip_num,
set_num,
species_code,
species_common_name,
haulbegin_day,
haulbegin_mon,
haulbegin_yr,
gender_code,
eye_fork
from newobs.lds_catch_v
where species_code = '305'
and eye_fork is not null
and eye_fork > '0'
order by haulbegin_yr"), HAULBEGIN_YR > 2003 | HAULBEGIN_YR==2003 & HAULBEGIN_MON > 7 )
ttr = proc.time() - p ## can use this to find out how long it took to download
# ## close connection
# close(con)

## combine the before & after catch data
catch1 = rbind(leg0,catch0)

## drop LL1065-LL1067 to avoid duplicates, per TM1602
catch = subset(catch1, !(TRIP_NUM %in% paste0('LL10',65:67)))

```
## Create a merge for each trip based on the set & trip number to avoid duplicates
```{r,warning = F, message = F}
## get only data since 2011 & drop data before Oct 2012
setgear = merge(set, gear, by = c('TRIP_NUM','SET_NUM'))
catch_merge0 = merge(catch0, setgear, by = c('TRIP_NUM','SET_NUM'))
catch_merge0$QTR = NA
catch_merge0$YRQTR = NA
catch_merge0$fqtr = NA
catch_merge = catch_merge0
## drop unneeded columns & order to match priorcatch
# names(priorcatch)
# sort(names(catch_merge0))
# catch_merge1 = catch_merge0[,c('TRIP_NUM','SET_NUM','TARGET_SPECIES_NAME',
#                         "SPECIES_COMMON_NAME","SPECIES_CODE","HAULBEGIN_YR",
#                         'HAULBEGIN_MON',"HAULBEGIN_DAY",'GENDER_CODE',"HKS_PER_FLT",
#                         'HAUL_BEGIN_LAT','HAUL_BEGIN_LON','EYE_FORK', 'QTR','YRQTR','fqtr')]
# 
# catch_merge = rbind(catch_merge1,priorcatch)
# catch_merge
```
## Extracting Quarter
I used calendar quarters for QTR and fishing year quarters for fQTR; seems to match what was done before. Then made a mixed variable using paste for year and calendar quarter.
```{r}
catch_merge$QTR = with(catch_merge, ifelse(HAULBEGIN_MON %in% 1:3,1,NA))
catch_merge$QTR = with(catch_merge, ifelse(HAULBEGIN_MON %in% 4:6,2,catch_merge$QTR))
catch_merge$QTR = with(catch_merge, ifelse(HAULBEGIN_MON %in% 7:9,3,catch_merge$QTR))
catch_merge$QTR = with(catch_merge, ifelse(HAULBEGIN_MON %in% 10:12,4,catch_merge$QTR))

catch_merge$YRQTR = paste0(catch_merge$HAULBEGIN_YR,".",catch_merge$QTR)

catch_merge$fqtr = with(catch_merge, ifelse(HAULBEGIN_MON %in% 1:3,3,NA))
catch_merge$fqtr = with(catch_merge, ifelse(HAULBEGIN_MON %in% 4:6,4,catch_merge$fqtr))
catch_merge$fqtr = with(catch_merge, ifelse(HAULBEGIN_MON %in% 7:9,1,catch_merge$fqtr))
catch_merge$fqtr = with(catch_merge, ifelse(HAULBEGIN_MON %in% 10:12,2,catch_merge$fqtr))


names(catch_merge)
catch_merge
```


```{r}
# write.csv(catch_merge,"G:/ISC BILLWG/sword_length_94_17", row.names = F)
```

