---
title: ""
format: 
 revealjs:
  theme: [default, customizations/presentation-custom.scss]
  logo: static/noaa-fisheries-logo.png
  css: customizations/logo.css
  slide-number: true
  revealjs-plugins:
  mermaid-format: svg
  cap-location: top
  smaller: true
embed-resources: true
execute:
    cache: true
---
# {.center}
::: {.titlebox style="text-align:center; font-size: 1.25em;"}

[Standardization of Pacific Blue Marlin Catch Per Unit Effort in the Hawaii Longline Fishery from 1995-2024]{.blue-title}

[Michelle Sculley, Ph.D.]{.blue style="font-size:0.5em;"}
<br>
[michelle.sculley@noaa.gov]{.blue style="font-size:0.5em;"}
:::

<!---This is how to add comments to a quarto file--->

## Pacific blue marlin in the - 

- Indo-Pacific blue marlin are caught by several fleets in the US: 
    - Hawaii longline (2 sectors targeting swordfish and tuna, respectively)
    - American Samoa longline
    - Handline
    - Troll
- Both HI LL and AS LL have observer program with varying amounts of coverage.
- BUM is a non-targeted but economically important species in the LL fisheries
- Overall catch appears to have increased in the last 10 years





## Data
:::{.r-fit-text}
- Data from 1995 - 2024 from the Pacific Islands Longline Logbook Program
- Deep Set :
    - Targets tuna
    - ≥ 10 hooks per float 1994-2003
    - ≥15 hooks per float after 2004
    - 3-10% observer coverage prior to 2004, 20% 2004 - 2020. 20-7% since 2020
    - accounts for 90% of the catch
- Shallow Set:
    - Targets Swordfish
    - <10 hooks per float 1994-2000
    - <15 hooks per float after 2004
    - Fishery closed from 2001-2004
    - 100% observer coverage after 2004
:::


## Blue Marlin Nominal CPUE
```{r}
source("BUM_dataexploration.r")

source("Plotting Pacific Map.R")

# set up the map data
BUMCPUE_Map <- aggregate(BUMCPUE$CPUE, by = list(BUMCPUE$Lat5, BUMCPUE$Lon5), mean)
names(BUMCPUE_Map) <- c("Lat", "Lon", "CPUE")
BUMCPUE_Map <- subset(BUMCPUE_Map, CPUE > 0)

grid_sf <- st_as_sf(
    BUMCPUE_Map,
    coords = c("Lon", "Lat"), # Specify which columns contain longitude and latitude
    crs = 4326 # Set the initial Coordinate Reference System (WGS84)
)

# Transform the 'sf' object to the target CRS (the one used for the map)
grid_sf_transformed <- grid_sf %>%
    st_transform(crs = target_crs)

Pacific_BaseMap +
    geom_sf(
        data = grid_sf_transformed,
        aes(color = CPUE),
        alpha = 0.7, size = 6
    ) +
    scale_color_viridis(discrete = FALSE) +
    guides(size = "none")


```




## Model Structure

- Delta-lognormal GLMM for both fleets combined
    - lognormal distribution for positive catches
    - binomial distribution with logit-link function for encounter probability
- Forward step-wise model selection on 13 potential explaintory variables
- Inclusion criteria were based upon an increase of percent deviance explained at least 0.10% and a Chi-squared likelihood ratio test



## Model Structure
:::: {.columns}

::: {.column width = "50%"}
- Operational variables:
    - Year
    - Month
    - Quarter
    - Hooks per float
    - bait type
    - Set time (Continuous and factor)
    - Set type
    - Vessel (random effect)
:::

:::{.column width = "50%"}
- Spatial and environmental
    - Latitude
    - Longitude
    - SST
    - PDO
    - SOI
    - ONI
:::

::::

## 
```{r}
#| layout-ncol: 2

SSTPlot

SOIPlot

PDOPlot

ONIPlot
```

##
```{r}
#| layout-ncol: 2

LatPlot

LonPlot
```

##
```{r}
#| layout-ncol: 2


BeginSetTimePlot

BeginPlot

HPFPlot

BaitPlot
```

## Final Models 
$$Probability of Encounter \sim Latitude + Year + Month + Longitude + Bait + HPF + Begin$$ 
9.7% Deviance Explained

$$log(CPUE) \sim Bait + Begin + Year + HPF + (1|Vessel)$$
10.2% Deviance Explained


## 
```{r}

CPUEPlot

```

## Diagnostics - Encounter Probability
```{r}
#| layout-ncol: 2

hist(residuals(BUM_PropPos, type = "pearson"), main="", xlab = "Pearson residuals")
qqnorm(residuals(BUM_PropPos), main="")
qqline(residuals(BUM_PropPos))
```

## Diagnostics - Positive Catches

```{r}
#| layout-ncol: 2

hist(residuals(BUM_Pos, type = "pearson"), main = "", xlab = "Pearson residuals")
qqnorm(residuals(BUM_Pos), main = "")
qqline(residuals(BUM_Pos))
```

## Partial Dependency Plots
```{r}
#| layout-ncol: 2


    pdp_prop_year
    pdp_prop_mon
    pdp_prop_lon
    pdp_prop_lat

```
Encounter Probability

## Partial Dependency Plots
```{r}
#| layout-ncol: 2

    pdp_prop_hpf
    pdp_prop_begin
    pdp_prop_bait


```
Encounter Probability

## Partial Dependency Plots
```{r}
#| layout-ncol: 2

    pdp_pos_year
    pdp_pos_bait
    pdp_pos_begin
    pdp_pos_hpf

```
Positivity Catches

## Residual plots - Encounter Probability
```{r}
#| layout-ncol: 2

    PPResid_Year
    PPResid_Month
PPResid_Lat
PPResid_Lon

```

## Residual plots - Encounter Probability

```{r}
#| layout-ncol: 2

    PPResid_HPF
    PPResid_Begin
    PPResid_Bait

```

## Residual plots - Positive catches

```{r}
#| layout-ncol: 2


   PosResid_Year
    PosResid_Bait
    PosResid_Begin
    PosResid_HPF

```