---
title: ""
format: 
 revealjs:
  theme: [default, customizations/presentation-custom.scss]
  logo: static/noaa-fisheries-logo.png
  css: customizations/logo.css
  slide-number: true
  revealjs-plugins:
  mermaid-format: svg
  cap-location: top
  smaller: true
embed-resources: true
execute: 
  cache: false
  fig-format: png
  fig-dpi: 300
  echo: false
  warning: false
  message: false
---
# {.center}
::: {.titlebox style="text-align:center; font-size: 1.25em;"}

[Summary of US Catch and Size of Pacific Blue Marlin]{.blue-title}

[Michelle Sculley, Ph.D.]{.blue style="font-size:0.5em;"}
<br>
[michelle.sculley@noaa.gov]{.blue style="font-size:0.5em;"}
:::


```{r}

#| echo: FALSE
source("c:/Users/Michelle.Sculley/Documents/2025_PacificBUM/Rscripts/CPUE standardization/CatchWP_Figures.r")
source("c:/Users/Michelle.Sculley/Documents/2025_PacificBUM/Rscripts/CPUE standardization/BUM_Length_Figures.R")
# print("Yes")
```



## Pacific Blue Marlin Annual Longline Catch
```{r}

AnnualLLCatch

```


## Pacific Blue Marlin Other Catch
```{r}
ggplot() +
geom_col(aes(x=YEAR, y=C_BLUE_MARLIN), data = subset(AnnualCatch, Source=="US Other Gears" & YEAR>2000)) +
    theme_bw() +
    scale_x_continuous(limits=c(2000, 2025), breaks=seq(2000,2025,5)) +
    scale_y_continuous(name="Catch (mt)")

```


## Blue Marlin Longline Catch
```{r}
library(ggplot2)
ggplot() +
    # World map polygons
    geom_polygon(
        data = world_df,
        aes(x = long, y = lat, group = group),
        fill = "grey80", color = "gray50"
    ) +

    # Plotting the effort data as points with binned sizes
    geom_point(
        data = HICatch_5yr,
        aes(x = Lon5, y = Lat5, size = Catch/1000),
        color = "red", alpha = 0.7
    ) +

    # Formatting X and Y axis labels with degrees
    scale_x_continuous(labels = function(x) paste0(abs(x), "°W")) +
    scale_y_continuous(labels = function(y) paste0(y, "°N")) +

    # Coordinate adjustments for map area
    coord_fixed(
        ratio = 1.3,
        xlim = c(-180, -100), # Adjust longitude range for Hawaii
        ylim = c(0, 40) # Adjust latitude range for Hawaii
    ) +

    # Customizing the legend
    guides(size = guide_legend(
        title = "Thousands of Fish Caught",
        override.aes = list(color = "red") # Ensure legend symbols are red
    )) +

    # Adding plot title and theme
    theme_minimal() +

    # Customizing the overall plot theme
    theme(
        legend.position = "bottom",
        legend.background = element_rect(fill = alpha("white", 0.7), color = "black", linewidth = 0.5),
        legend.title = element_text(size = 10, face = "bold"),
        legend.text = element_text(size = 9),
        plot.title = element_text(size = 14, face = "bold"),
        plot.margin = margin(20, 20, 20, 20),
        panel.grid.major = element_line(color = "gray70"),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
        axis.text.x = element_text(size = 9),
        axis.text.y = element_text(size = 9),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.background = element_rect(fill = "white", color = NA) # Set white background
    ) + 
    facet_wrap(~YrGroup, ncol=3, labeller = labeller(YrGroup=YearLabels)) 

```



## Blue marlin Longline Catch by Quarter
```{r}

CatchMapQTR

```


## Blue marlin HI LL Catch by Set
```{r}

CatchMapSet


```


## Blue marlin HI LL effort by Set
```{r}
EffortMapSet

```

## HI LL Annual Mean Length 
```{r}

q <- boxplot(Length ~ Year, data = subset(BUM_Length, Fleet=="LL"),xlab = "Year", ylab = "Eye-Fork Length (cm)", ylim = c(0, 350))
text(x = 1:31, y = 10, labels = q$n, cex = 0.75)

```

## Annual Mean length by fleet
```{r}

AnnualMeanLine

```


## American Samoa LL Annual Mean Length
```{r}

z <- boxplot(Length ~ Year, data = subset(BUM_Length, Fleet=="AS"),xlab = "Year", ylab = "Eye-Fork Length (cm)", ylim = c(0, 350))
text(x = 1:19, y = 10, labels = z$n, cex = 0.75)

```

## Histogram of HI LL lengths overall and by quarter

```{r}
#| layout-ncol: 2
#| echo: false
#| warning: false
#| message: false

ggplot() +
    geom_histogram(aes(x = Length), data = subset(BUM_Length, Fleet=="LL"), bins = 50, fill = "grey90", color = "black") +
    theme_bw()

ggplot() +
    geom_histogram(aes(x = Length), data = subset(BUM_Length, Fleet=="LL"), bins = 50, color = "black", fill = "grey90") +
    theme(
        panel.background = element_blank(), panel.grid.major = element_line(colour = "grey90"),
        panel.grid.minor = element_blank(),
        strip.background = element_blank(),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        strip.text = element_text(size = 16)
    ) +
    geom_vline(aes(xintercept = 177.76), data = data.frame(1), color = "red", linetype = "dashed") +
    facet_wrap(~Quarter) 


```




## Mean LL length by area
```{r}
source("Plotting Pacific Map.R")

# set up the map data

grid_sf <- st_as_sf(
    SLen5,
    coords = c("Lon", "Lat"), # Specify which columns contain longitude and latitude
    crs = 4326 # Set the initial Coordinate Reference System (WGS84)
)

# Transform the 'sf' object to the target CRS (the one used for the map)
grid_sf_transformed <- grid_sf %>%
    st_transform(crs = target_crs)

Pacific_BaseMap +
    geom_sf(
        data = grid_sf_transformed,
        aes(color = Length),
        alpha = 0.7, size=6
    ) +
    scale_color_viridis(discrete = FALSE) +
    guides(size = "none") +
    theme(legend.position = "bottom")

```

## Mean LL length by quarter

```{r}

grid_sf <- st_as_sf(
    SLenQ5,
    coords = c("Lon", "Lat"), # Specify which columns contain longitude and latitude
    crs = 4326 # Set the initial Coordinate Reference System (WGS84)
)

# Transform the 'sf' object to the target CRS (the one used for the map)
grid_sf_transformed <- grid_sf %>%
    st_transform(crs = target_crs)

Pacific_BaseMap +
    geom_sf(
        data = grid_sf_transformed,
        aes(color = Length),
        alpha = 0.7, size=3
    ) +
    scale_color_viridis(discrete = FALSE) +
    guides(size = "none") +
    facet_wrap(~Quarter)+
    theme(legend.position = "bottom")


```